Randomly-Scoped Lambda Calculus

### 2013-01-27 RSLC/paper/paper.tex
minor changes to rslc paper
rslc typoes

However, this isn't really ``programming'', and Sully thought for sure I wouldn't be able to do anything useful with this language, which of course I wasn't going to stand by. The first thing I wanted to do was to recursively generate a random natural number, something like this:
\begin{equation}
\label{eq:randomnat}
\Y~(\lambda f. \lambda n. (\lambda x. \lambda x. x)~n~(f~\Suc{n}))~\Zero
\end{equation}
This term would randomly choose whether to output $n$, its argument, or to call itself with $\Suc{n}$. However, the conventional Y-combinator doesn't work as intended in RSLC, because once a function is substituted into itself for its first argument, subsequent arguments will be captured in the substituted version.
Even ignoring the internal workings of the combinator, and supposing $\Y$ satisfies $\Y~g \mapsto^* g~(\Y~g)$, here's what happens:

@@

The variable $\thing{\bf t}$ accumulates ``$\lambda x$''s as $\rho$ recurses (in the $\Suc{n_2}$ case). You might ask: why does the $\lambda f$ need to be inside the $\mathsf{case}$ statement?
Because otherwise in {\sc Eval-Case-2}$n_2$ would get captured in the thing substituted for $f$.

@@

Got here. Up until now I showed how random scope adds inherent nondeterminism that can be harnessed to simulate dice anda make silly graphs. The other thing Sully didn't believe would be possible was writing deterministic RSLC programs that {\em avoid} using ``the feature''.

@@

\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
To compute fibonacci numbers, I used the standard Church encoding for pairs:
\begin{eqnarray*}
(e_1, e_2) &=& \lambda b.~b~e_1~e_2 \\
\fst~e &=& e (\lambda x.\lambda y. x) \\
\snd~e &=& e (\lambda x.\lambda y. y) \\
\end{eqnarray*}
Hence:
\newcommand{\fib}{\Phi}
\newcommand{\fibxa}{\tilde{x}_1}
\newcommand{\fibxb}{\tilde{x}_2}
\newcommand{\fibn}{\tilde{n}}
\newcommand{\fibf}{\tilde{f}}
\newcommand{\fibp}{\tilde{p}}
\newcommand{\fibnn}{\tilde{n}'}
\begin{center}
\begin{tabular}{rcl}
$\fib$ & = &
\hspace{-1em}\multirow{6}{*}{
\begin{tabular}{lllll}
\multicolumn{5}{l}{$\lambda \fibn.$} \\
& \multicolumn{4}{l}{$\mathsf{case}~\fibn~\mathsf{of}$} \\
&& $\Zero$ & $\Rightarrow$ & $\lambda \fibf.~(\Zero,~\Suc{\Zero})$ \\
&& $\Suc{\fibnn}$ &$\Rightarrow$& $\lambda \fibf.$ \\
&&
\multicolumn{3}{l}{
\begin{tabular}{lll}
& \textsf{let} & $\fibp~=~\Ya~\fibnn~\fibf$ \\
& & $\fibxa~=~\fst~\fibp$ \\
& & $\fibxb~=~\snd~\fibp$ \\
& \textsf{in} & $(\fibxb,~\mathsf{add}~\fibxa~\fibxb)$
\end{tabular}
}
\end{tabular}
}
\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
$\mathsf{fib}~N$ &=& $\Ya~N~\fib$
\end{tabular}
\end{center}

>>>>

However, this isn't really ``programming'', and Sully thought for sure I wouldn't be able to do anything useful with this language, which of course I wasn't going to stand by. The first thing I wanted to do was to recursively generate a random natural number, something like this:
\begin{equation}
\label{eq:randomnat}
\Y~(\lambda f. \lambda n. (\lambda x. \lambda x. x)~n~(f~\Suc{n}))~\Zero
\end{equation}
This term would randomly choose whether to output $n$, its argument, or to call itself with $\Suc{n}$. However, the conventional Y-combinator doesn't work as intended in RSLC, because once a function is substituted into itself for its first argument, subsequent arguments will be captured in the substituted version.
Even ignoring the internal workings of the combinator, and supposing some $\Y$ that satisfies $\Y~g \mapsto^* g~(\Y~g)$, here's what happens:

@@

The variable $\thing{\bf t}$ accumulates ``$\lambda x$''s as $\rho$ recurses (in the $\Suc{n_2}$ case). You might ask: why does the $\lambda f$ need to be inside the $\mathsf{case}$ statement?
Because otherwise, in {\sc Eval-Case-2}, $n_2$ would get captured in the thing substituted for $f$.

@@

Got here. Up until now I showed how random scope adds inherent nondeterminism that can be harnessed to simulate dice and make silly graphs. The other thing Sully didn't believe would be possible was writing deterministic RSLC programs that {\em avoid} using ``the feature''.

@@

\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
To compute fibonacci numbers, I used the standard Church encoding for pairs:
\begin{eqnarray*}
(e_1, e_2) &=& \lambda b.~b~e_1~e_2 \\
\fst~e &=& e (\lambda x.\lambda y. x) \\
\snd~e &=& e (\lambda x.\lambda y. y) \\
\end{eqnarray*}
Hence:
\newcommand{\fib}{\Phi}
\newcommand{\fibxa}{\tilde{x}_1}
\newcommand{\fibxb}{\tilde{x}_2}
\newcommand{\fibn}{\tilde{n}}
\newcommand{\fibf}{\tilde{f}}
\newcommand{\fibp}{\tilde{p}}
\newcommand{\fibnn}{\tilde{n}'}
\begin{center}
\begin{tabular}{rcl}
$\fib$ & = &
\hspace{-1em}\multirow{6}{*}{
\begin{tabular}{lllll}
\multicolumn{5}{l}{$\lambda \fibn.$} \\
& \multicolumn{4}{l}{$\mathsf{case}~\fibn~\mathsf{of}$} \\
&& $\Zero$ & $\Rightarrow$ & $\lambda \fibf.~(\Zero,~\Suc{\Zero})$ \\
&& $\Suc{\fibnn}$ &$\Rightarrow$& $\lambda \fibf.$ \\
&&
\multicolumn{3}{l}{
\begin{tabular}{lll}
& \textsf{let} & $\fibp~~=~\Ya~\fibnn~\fibf$ \\
& & $\fibxa~=~\fst~\fibp$ \\
& & $\fibxb~=~\snd~\fibp$ \\
& \textsf{in} & $(\fibxb,~\mathsf{add}~\fibxa~\fibxb)$
\end{tabular}
}
\end{tabular}
}
\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
&&\\
$\mathsf{fib}~N$ &=& $\Ya~N~\fib$
\end{tabular}
\end{center}
